Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and neurodevelopment: A study based on the parametric g-formula

### 2024-02-06 output/paper/manuscript.qmd
Modified almost all text based on MV. Added authors and roles.

\acr{opps} are one group of \acr{edcs} that may have a deleterious effect on neurodevelopment, although the few studies assessing exposure during childhood and through the use of biomarkers suffered from a series of limitations, including a small sample size and few \acr{opp} metabolites measured [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Exposure to phthalates and their metabolites has been associated to several adverse neurodevelopmental outcomes, but these studies were limited to few phthalate metabolites and small study populations [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. The effects of exposure to \acr{bpa} during childhood on cognitive functions are still unclear [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

Moreover, little is known about the biological mechanisms of action [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].
There is also evidence that exposure to certain \acr{edcs}, specifically phthalates, might interfere with the \acr{hpa} axis and might interact with the glucocorticoid receptor [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023].

The HPA axis, which can be activated by stress, is responsible for the production of glucocorticosteroids. The brain, and its proper functioning, is a potential target, due to the presence of receptors for these hormones [@LupienMcEwenGunnar:2009; @SunLiJin:2018]. Glucocorticosteroids are necessary for brain maturation, although their under- or over-production might interfere with its normal development and ultimately lead to long-term impaired functioning [@LupienMcEwenGunnar:2009; @SearsLiuLanphear:2023].

Taken together, these results suggest that the negative influence of exposure to certain \acr{edcs} on neurodevelopmental outcomes, might be mediated, at least partially, by disruption of the \acr{hpa} axis' homeostasis.
In the present study, we thus estimated associations between 1) non-persistent \acr{edcs} and attention, 2) non-persistent \acr{edcs} and glucocorticosteroids, and 3) glucocorticosteroids and attention, using the parametric g-formula and marginal contrasts, in children of a large prospective birth cohort in Europe.

@@

The \acr{helix} is an ongoing project which aims to characterize early-life exposures and their potential association with endogenous biomarkers and health outcomes [@VrijheidSlamaRobinson:2014]. It consists of six existing population-based birth cohort studies across Europe: BiB (Born in Bradford, UK) [@WrightSmallRaynor:2013], EDEN (Study of determinants of pre- and postnatal developmental, France) [@HeudeForhanSlama:2016], INMA (Environment and Childhood, Spain) [@GuxensBallesterEspada:2012], KANC (Kaunas Cohort, Lithuania) [@GrazulevicieneDanileviciuteNadisauskiene:2009], MoBa (The Norwegian Mother and Child Cohort Study, Norway) [@MagnusIrgensHaug:2006], and Rhea (Mother–Child Cohort in Crete, Greece) [@ChatziPlanaDaraki:2009], for a total of 32,000 mother-child pairs. A \acr{helix} subcohort of 1,200 mother-child pairs was fully characterized for the external and internal exposome, including exposure and omics biomarkers during childhood. Eligibility criteria for inclusion in the \acr{helix} subcohort included: a) age 6-11 years, with a preference for 7-9 years; b) availability of sufficient stored pregnancy blood and urine samples; c) availability of complete address history from first to last follow-up; d) no serious health problems, which might affect the results of the clinical testing. Further information can be found in @MaitreBontCasas:2018.

Ethical permission was obtained from the relevant authorities in the corresponding country.

@@

For each research question, defined by a specific type of exposure and outcome, the minimal set of covariates for inclusion in the analyses was selected on the basis of a \acr{dag} built with DAGitty [@TextorvanderZanderGilthorpe:2016] and ggdag [@Barrett:2023]. The sets of covariates were selected to estimate the total effect of the exposure on the outcome. Further, each minimal adjustment set was *augmented* with precision covariates, defined as the set of parents variable of the outcome that are not parents of the exposure. The adjustment sets are provided in the Supplementary Material as text files compatible with DAGitty.

@@

```{r}
info_edcs <- myphd::edcs_information()
print_tidy_abbreviations_edcs <- function(df, class_) {
  tmp <- df |>
    dplyr::filter(class == class_) |>
    dplyr::select(chem_id, full_name) |>
    dplyr::arrange(full_name) |>
    dplyr::mutate(
      ret = glue::glue("{full_name} ({chem_id})")
    )
  tmp$ret |>
    paste0(collapse = ", ")
}
```

Children were assessed between December 2013 and February 2016, and included neurological testing and urine collection. Urine samples of the night before and the first morning void on the day of the visit were combined to provide a more reliable exposure assessment. Non-persistent \acr{edcs} assessed in urine samples from children included phthalate metabolites, phenols, and \acr{op} pesticide metabolites. A list of the environmental chemicals determined in urine samples and used for the present study is given in @supptbl-info-chems. Briefly, we analyzed a total of 7 phenols (`r print_tidy_abbreviations_edcs(info_edcs, "Phenols")`), 6 non-specific organophosphate pesticide metabolites (`r print_tidy_abbreviations_edcs(info_edcs, "OP pesticide metabolites")`), and 10 phthalate metabolites (`r print_tidy_abbreviations_edcs(info_edcs, "Phthalate metabolites")`) originating from 6 distinct phthalate parent compounds.
The laboratory protocols for the analysis are described elsewhere [@HaugSakhiCequier:2018].

@@

To assess the levels of glucocorticosteroids and their metabolites, LC-MS/MS analysis was applied at the Applied Metabolomics Research Group, IMIM (Hospital del Mar Medical Research Institute). The laboratory protocols for the analysis are described elsewhere [@MarcosRenauCasals:2014; @Gomez-GomezPozo:2020]. Of the 1,004 urine samples, 980 children were matched to the \acr{helix} subcohort.

@@

Neurodevelopmental outcomes were assessed with standardized, non-linguistic, and culturally blind computer tests, including the \acr{ant} [@RuedaFanMcCandliss:2004]. Further information can be found in @MaitreBontCasas:2018. Briefly, it is a computerized test that provides a measure of efficiency in three different functions of attention: alerting, orienting, and executive attention. The outcome of interest for the present study is the \acr{hitrtse} [@SunyerEsnaolaAlvarez-Pedrerol:2015], a measure of response speed consistency throughout the test. A high \acr{hitrtse} indicates highly variable reactions, and is considered a measure of inattentiveness.

@@

Missing values in the clinical outcome were imputed similarly using kNN. Similarly, we natural log-transformed them to improve model fit, assessed with posterior predictive checks. Values of the clinical outcome were expressed in \acr{mss}.

@@

Stabilized balancing weights were estimated using the energy method available in the `WeightIt` R package [@Greifer:2023]. This methods estimates weights by minimizing an energy statistic related to covariate balance [@HulingGreiferChen:2023], thus avoiding the need to specify a parametric model.
Weights below the 0.1 and above the 0.9 quantiles were trimmed. Trimming might lead to decreased covariate balance and potentially change the estimand, but can also decrease the variability of the weights.
Covariate balance was assessed using functionalities provided by the `cobalt` R package [@Greifer:2023a]. Specifically, we used *Love* plots to visualize covariate balance before and after adjusting.

@@

## Participants {#sec-res-participants}

In total, 1,297 children were included in the dataset. Of these, 976 had information also about the glucocorticosteroids. The sample consisted of 55% girls. The median \acr{hitrtse} was 300 ms (\acr{iqr}, 231-368), with lower median values for EDEN, MOBA, and SAB, corresponding to the cohorts with older children. At the time of visit, the median age of the children was 8.06 years. The children were mostly Caucasian (90%) or Pakistani (6.2%). @tbl-pop-desc provides descriptive statistics for the outcome and covariates, by cohort and overall. Codebooks for the used covariates, by research question, are provided in @supptbl-codebook-1, @supptbl-codebook-2, and @supptbl-codebook-3.

## Descriptive data {#sec-res-pop-desc}

@@

## Main results {#sec-res-main}

The effective sample sizes before and after balancing weights estimation are presented in Tables [-@supptbl-balance-1; -@supptbl-balance-2; -@supptbl-balance-3], while basic summary statistics of the estimated balancing weights are presented in Tables [-@supptbl-weights-1; -@supptbl-weights-2; -@supptbl-weights-3]. As expected, the median value of the weights for each exposure was close to $1.00$.

@Fig-marginal-1 presents the forest plot for the marginal contrasts of the non-persistent \acr{edcs} on \acr{hitrtse}. Overall, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive marginal contrast, indicating an increase in the values of \acr{hitrtse}. Most of the \acr{ci}s included the null effect, though. Statistically significant effects were observed for the phenol MEPA (`r format_res(meffects1, "mepa", "hitrtse")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "hitrtse")`), oh-MiNP (`r format_res(meffects1, "ohminp", "hitrtse")`), and MEHP (`r format_res(meffects1, "mehp", "hitrtse")`). The \acr{opp} DETP was strongly negatively associated with \acr{hitrtse} (`r format_res(meffects1, "detp", "hitrtse")`).

@Fig-marginal-2 presents the forest plot for the marginal contrasts of the non-persistent \acr{edcs} on cortisone production, cortisol production, and corticosterone production. Overall, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive marginal contrast, indicating an increase in the total production of these metabolites, with the exception of BUPA, which was associated with negative marginal contrasts for all three outcomes, and MiBP, which was associated with a negative marginal contrast for total cortisone production only.
The majority of the effects for the phenols and phthalate metabolites included the null. The phenol BPA showed the largest marginal contrasts across all three outcomes (cortisone production, `r format_res(meffects2, "bpa", "cortisone production")`; cortisol production, `r format_res(meffects1, "bpa", "cortisol production")`; corticosterone production, `r format_res(meffects1, "bpa", "corticosterone production")`).

@@

## Other analyses {#sec-res-other}

### Effect modification by sex

@@

Basic summary statistics of the estimated balancing weights for effect modification are presented in Tables [-@supptbl-weights-1sa; -@supptbl-weights-2sa; -@supptbl-weights-3sa]. As expected, the median value of the weights for each exposure was close to $1.00$.

@@

### Sensitivity to exposure assessment

@@

The impact of exposure to (non-persistent) \acr{edcs} on human health has attracted considerable research interest. While research in this area has mainly investigated the effects of prenatal exposure on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022], little is still known about later-in-life exposure and childhood neurodevelopment. In this study, consisting of almost 1,000 children from 6 European birth cohorts, we observed that short-term childhood exposure to non-persistent \acr{edcs} had negative effects on \acr{hitrtse} and total production of cortisol, cortisone, and corticosterone, with the majority of the \acr{ci}s including the null effect. Increased production of these glucocorticosteroids did not seem to affect \acr{hitrtse}. Some of these effects differed for females and males, including significant differences for the effects of increased production of cortisone and corticosterone on \acr{hitrtse}. Specifically, an increased production of these glucocorticosteroids was associated with lower values of \acr{hitrtse} for females, and higher values for males. Taken together, these results suggest that these non-persistent \acr{edcs} might be responsible for perturbations of the \acr{hpa} axis' homeostasis, and that higher levels of these glucocorticosteroids might interfere with different functions of attention in a sex-specific manner.

We are not aware of prior studies specifically investigating the effects of exposure to \acr{edcs} in relation to \acr{hitrtse}. The literature on \acr{edcs} and neurodevelopment in children has mostly focused on \acr{opps} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023], and \acr{bpa} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019], in relation to \acr{adhd} [@BouchardBellingerWright:2010; @TewarAuingerBraun:2016; @YuDuChiou:2016; @LiZhangKuang:2018], and intelligence scales [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023]. Few studies have looked into different classes of \acr{edcs} [@ShoaffCoullWeuve:2020 in relation with the Conners Attention Deficit Scale and the Behavior Assessment System for Children; @OhKimKannan:2023 in relation with \acr{adhd} symptoms]. Overall, and consistent with our results, these studies seem to provide further evidence of the negative effects of several \acr{edcs} on neurodevelopment in children. While not all these studies have investigated effect modification by sex, it seems that these negative effects are stronger in males. A major limitation of these studies is the reliance on spot urine samples, that might not be representative of long-term exposures.

Our results are consistent with prior research that associated exposure to certain \acr{edcs} with higher levels of cortisol [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023]. There are some differences, though. First, these studies only focus on phthalates, either as individual metabolites or as mixture. Second, exposure assessment in @SunLiJin:2018 and @KimLeeMoon:2018 was performed during gestation or the first 15 months of life, respectively. Finally, the glucocorticosteroids were measured in cord blood [@SunLiJin:2018] and hair [@SearsLiuLanphear:2023]. Contrary to these studies, we did find effect modification by sex. We are not aware of other epidemiological studies investigating phthalates metabolites, phenols, and \acr{opps}, in relation to urinary glucocorticosteroids in childhood.

We are also not aware of prior studies specifically investigating the effects of elevated levels of glucocorticosteroids in relation to \acr{hitrtse}, although there is evidence that under- or over-production of glucocorticosteroids interfere with the normal development of the brain [@LupienMcEwenGunnar:2009]. While we did find sex-specific evidence of an effect, their clinical relevance is questionable.

These findings should be interpreted in light of the following limitations and strengths.

Limitations include the cross-sectional design of the present study. Importantly, the non-persistent \acr{edcs} were measured in a pool of night and morning urine samples before the clinical visit, whereas the glucocorticosteroids were measured in the night urine sample only. Although we included a wide range of confounders there is the possibility, as with other observational studies, of residual confounding, which might lead to a bias away from the null. Some of the confounders indicated in the adjustment sets had to be remove due to large fractions of missing values. There is further the possibility of misspecification of the outcome model, although we included a spline of the exposure to relax some of the linearity assumptions. The use of more data-adaptive learners was excluded due to the relatively small sample size. We finally acknowledge the possibility that some of chemicals might not act independently (mixture effect). Further research is thus warranted.

@@

In conclusion, in a study of almost 1,000 children from 6 European birth cohorts, we observed that (i) exposure to non-persistent \acr{edcs} in childhood might have short-term effects on \acr{hitrtse} in childhood, (ii) exposure to non-persistent \acr{edcs} in childhood might disrupt the \acr{hpa} axis in childhood, and (iii) disruption of the \acr{hpa} axis in childhood might have short-term, sex-specific effects on \acr{hitrtse}. Future studies should investigate how glucocorticosteroids might mediate the adverse effects of exposure to \acr{edcs} on childhood neurodevelopment in larger populations.

>>>>

\acr{opps} are one group of \acr{edcs} that may have a deleterious effect on neurodevelopment, although the few studies assessing exposure during childhood and through the use of biomarkers suffered from a series of limitations, including a small sample size and few \acr{opp} metabolites measured [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Exposure to phthalates and their metabolites during childhood and early adolescence has been associated to several adverse neurodevelopmental outcomes, but these studies were limited to few phthalate metabolites and small study populations [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. The effects of exposure to \acr{bpa} during childhood on cognitive functions are still unclear [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

Moreover, little is known about the biological mechanisms of action [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. There is some toxicological evidence, however, that exposure to certain \acr{edcs}, specifically phthalates, might interfere with the \acr{hpa} axis and might interact with the glucocorticoid receptor [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023]. The HPA axis, which can be activated by stress, is responsible for the production of glucocorticosteroids. The brain, and its proper functioning, is a potential target, due to the presence of receptors for these hormones [@LupienMcEwenGunnar:2009; @SunLiJin:2018]. Glucocorticosteroids are necessary for brain maturation, although their under- or over-production might interfere with its normal development and ultimately lead to long-term impaired functioning [@LupienMcEwenGunnar:2009; @SearsLiuLanphear:2023].

Taken together, these results suggest that the negative influence of exposure to certain \acr{edcs} on neurodevelopmental outcomes, might be mediated, at least partially, by disruption of the \acr{hpa} axis' homeostasis.
In the present study, we thus estimated associations between 1) non-persistent \acr{edcs} and attention, 2) non-persistent \acr{edcs} and glucocorticosteroids, and 3) glucocorticosteroids and attention, using the parametric g-formula and marginal contrasts, in children of a large cohort in Europe.

@@

The \acr{helix} project aims to characterize early-life exposures and their potential association with endogenous biomarkers and health outcomes [@VrijheidSlamaRobinson:2014]. It consists of six existing population-based birth cohort studies across Europe: BiB (Born in Bradford, UK) [@WrightSmallRaynor:2013], EDEN (Study of determinants of pre- and postnatal developmental, France) [@HeudeForhanSlama:2016], INMA (Environment and Childhood, Spain) [@GuxensBallesterEspada:2012], KANC (Kaunas Cohort, Lithuania) [@GrazulevicieneDanileviciuteNadisauskiene:2009], MoBa (The Norwegian Mother and Child Cohort Study, Norway) [@MagnusIrgensHaug:2006], and Rhea (Mother–Child Cohort in Crete, Greece) [@ChatziPlanaDaraki:2009]. The \acr{helix} subcohort of 1,301 mother-child pairs was fully characterized for the external and internal exposome, including exposure and omics biomarkers during childhood [@MaitreBontCasas:2018]. Eligibility criteria for inclusion in the \acr{helix} subcohort included: a) age 6-11 years, with a preference for 7-9 years; b) availability of sufficient stored pregnancy blood and urine samples; c) availability of complete address history from first to last follow-up; d) no serious health problems, which might affect the results of the clinical testing. Ethical permission was obtained from the relevant authorities in the corresponding country.

@@

@@

```{r}
info_edcs <- myphd::edcs_information()
print_tidy_abbreviations_edcs <- function(df, class_) {
  tmp <- df |>
    dplyr::filter(class == class_) |>
    dplyr::select(short_name, full_name) |>
    dplyr::arrange(full_name) |>
    dplyr::mutate(
      ret = glue::glue("{full_name} ({short_name})")
    )
  tmp$ret |>
    paste0(collapse = ", ")
}
```

Children were assessed between December 2013 and February 2016, and assessments included neurological testing and urine collection. Urine samples of the night before and the first morning void on the day of the visit were combined to provide a more reliable exposure assessment. Non-persistent \acr{edcs} assessed in the urine samples included phthalate metabolites, phenols, and \acr{op} pesticide metabolites. A list of the environmental chemicals determined in urine samples and used for the present study is given in @supptbl-info-chems. Briefly, we analyzed a total of 7 phenols (`r print_tidy_abbreviations_edcs(info_edcs, "Phenols")`), 6 non-specific organophosphate pesticide metabolites (`r print_tidy_abbreviations_edcs(info_edcs, "OP pesticide metabolites")`), and 10 phthalate metabolites (`r print_tidy_abbreviations_edcs(info_edcs, "Phthalate metabolites")`) originating from 6 distinct phthalate parent compounds.
The laboratory protocols for the analysis are described elsewhere [@HaugSakhiCequier:2018].

@@

To assess the levels of glucocorticosteroids and their metabolites, LC-MS/MS analysis was applied at the Applied Metabolomics Research Group, IMIM (Hospital del Mar Medical Research Institute). The laboratory protocols for the analysis are described elsewhere [@MarcosRenauCasals:2014; @Gomez-GomezPozo:2020]. Measurements were available for 980 children of the \acr{helix} subcohort.

@@

Neurodevelopmental outcomes were assessed with standardized, non-linguistic, and culturally blind computer tests, including the \acr{ant} [@RuedaFanMcCandliss:2004], which provides a measure of efficiency in three different functions of attention: alerting, orienting, and executive attention. The tests were administered in a standardized way, and with minimal interference from the field workers. Further information can be found in @MaitreBontCasas:2018. The outcome of interest for the present study is the \acr{hitrtse} [@SunyerEsnaolaAlvarez-Pedrerol:2015], a measure of response speed consistency throughout the test. A high \acr{hitrtse} indicates highly variable reactions, and is considered a measure of inattentiveness.

For each research question, defined by a specific type of exposure and outcome, the minimal set of covariates for inclusion in the analyses was selected on the basis of a \acr{dag} built with DAGitty [@TextorvanderZanderGilthorpe:2016] and ggdag [@Barrett:2023]. The sets of covariates were selected to estimate the total effect of the exposure on the outcome. Further, each minimal adjustment set was *augmented* with precision covariates, defined as the set of parents variable of the outcome that are not parents of the exposure. The adjustment sets are provided in the Supplementary Material as text files compatible with DAGitty. Codebooks for the used covariates, by research question, are provided in @supptbl-codebook-1, @supptbl-codebook-2, and @supptbl-codebook-3.

@@

Missing values in the clinical outcome were imputed similarly using kNN. We natural log-transformed these to improve model fit, assessed with posterior predictive checks. Values of the clinical outcome were expressed in \acr{mss}.

@@

To reduce the effect of measured confounders on the exposure-outcome association, stabilized balancing weights were estimated using the energy method available in the `WeightIt` R package [@Greifer:2023]. This methods estimates weights by minimizing an energy statistic related to covariate balance [@HulingGreiferChen:2023], thus avoiding the need to specify a parametric model.
Weights below the 0.1 and above the 0.9 quantiles were trimmed. Trimming might lead to decreased covariate balance and potentially change the estimand, but can also decrease the variability of the weights.
Covariate balance was assessed using functionalities provided by the `cobalt` R package [@Greifer:2023a]. Specifically, we used *Love* plots to visualize covariate balance before and after adjusting.

@@

In total, 1,297 children were included in the study population. Of these, 976 had information also about the glucocorticosteroids. The sample consisted of 55% girls. The median \acr{hitrtse} was 300 ms (\acr{iqr}, 231-368), with lower median values for EDEN, MOBA, and INMA, corresponding to the cohorts with older children. At the time of visit, the median age of the children was 8.06 years. The children were mostly Caucasian (90%), and the largest minority were of Pakistani origin (6.2%). @tbl-pop-desc provides descriptive statistics for the outcome and covariates, by cohort and overall.

@@

The effective sample sizes before and after balancing weights estimation are presented in Supplementary Tables [-@supptbl-balance-1; -@supptbl-balance-2; -@supptbl-balance-3], while basic summary statistics of the estimated balancing weights are presented in Supplementary Tables [-@supptbl-weights-1; -@supptbl-weights-2; -@supptbl-weights-3]. As expected, the median value of the weights for each exposure was close to $1.00$.

@Fig-marginal-1 presents the forest plot for the marginal contrasts of the non-persistent \acr{edcs} on \acr{hitrtse}. For most \acr{edcs}, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive marginal contrast, indicating an increase in the values of \acr{hitrtse}. Most of the \acr{ci}s included the null effect, though. Statistically significant effects were observed for the phenol MEPA (`r format_res(meffects1, "mepa", "hitrtse")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "hitrtse")`), oh-MiNP (`r format_res(meffects1, "ohminp", "hitrtse")`), and MEHP (`r format_res(meffects1, "mehp", "hitrtse")`). The \acr{opp} DETP was strongly negatively associated with \acr{hitrtse} (`r format_res(meffects1, "detp", "hitrtse")`).

@Fig-marginal-2 presents the forest plot for the marginal contrasts of the non-persistent \acr{edcs} on cortisone production, cortisol production, and corticosterone production. For most \acr{edcs}, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive marginal contrast, indicating an increase in the total production of these metabolites. Exceptions were BUPA, which was associated with negative marginal contrasts for all three outcomes, and MiBP, which was associated with a negative marginal contrast for total cortisone production only.
The majority of the effects for the phenols and phthalate metabolites included the null. The phenol BPA showed the largest marginal contrasts across all three outcomes (cortisone production, `r format_res(meffects2, "bpa", "cortisone production")`; cortisol production, `r format_res(meffects1, "bpa", "cortisol production")`; corticosterone production, `r format_res(meffects1, "bpa", "corticosterone production")`).

@@

## Effect modification by sex

@@

Basic summary statistics of the estimated balancing weights for effect modification are presented in Supplementary Tables [-@supptbl-weights-1sa; -@supptbl-weights-2sa; -@supptbl-weights-3sa]. As expected, the median value of the weights for each exposure was close to $1.00$.

@@

@@

The impact of exposure to \acr{edcs} on human health has attracted considerable research interest. While research in this area has mainly investigated the effects of prenatal exposure on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022], little is still known about childhood exposure. In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to non-persistent \acr{edcs} had negative effects on \acr{hitrtse} and total production of cortisol, cortisone, and corticosterone, although the majority of the \acr{ci}s included the null effect. Increased production of these glucocorticosteroids did not seem to affect \acr{hitrtse}. Some of these effects differed for females and males, including significant differences for the effects of increased production of cortisone and corticosterone on \acr{hitrtse}. Specifically, an increased production of these glucocorticosteroids was associated with lower values of \acr{hitrtse} for females, and higher values for males. Taken together, these results suggest that these non-persistent \acr{edcs} might be responsible for perturbations of the \acr{hpa} axis' homeostasis, and that higher levels of these glucocorticosteroids might interfere with different functions of attention in a sex-specific manner.

We are not aware of prior studies specifically investigating the effects of exposure to \acr{edcs} in relation to \acr{hitrtse}. The literature on \acr{edcs} and neurodevelopment in children has mostly focused on \acr{opps} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023], and \acr{bpa} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019], in relation to \acr{adhd} [@BouchardBellingerWright:2010; @TewarAuingerBraun:2016; @YuDuChiou:2016; @LiZhangKuang:2018], and intelligence scales [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023]. Few studies have looked into different classes of \acr{edcs} [@ShoaffCoullWeuve:2020 in relation with the Conners Attention Deficit Scale and the Behavior Assessment System for Children; @OhKimKannan:2023 in relation with \acr{adhd} symptoms]. Overall, and consistent with our results, these studies seem to provide further evidence of the adverse effects of several \acr{edcs} on neurodevelopment in children. While not all these studies have investigated effect modification by sex, it seems that these adverse effects are stronger in males. A major limitation of these studies is the reliance on spot urine samples, that might not be representative of long-term exposures.

Our results are consistent with prior epidemiological research that associated exposure to certain \acr{edcs} with higher levels of cortisol [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023]. There are some differences, though. First, these studies only focus on phthalates, either as individual metabolites or as mixture. Second, exposure assessment in @SunLiJin:2018 and @KimLeeMoon:2018 was performed during gestation or the first 15 months of life, respectively. Finally, the glucocorticosteroids were measured in cord blood [@SunLiJin:2018] and hair [@SearsLiuLanphear:2023]. Contrary to these studies, we did find effect modification by sex. We are not aware of other epidemiological studies investigating phthalates metabolites, phenols, and \acr{opps}, in relation to urinary glucocorticosteroids in childhood.

We are also not aware of prior epidemiological studies specifically investigating the effects of elevated levels of glucocorticosteroids in relation to \acr{hitrtse}, although there is evidence that under- or over-production of glucocorticosteroids interfere with the normal development of the brain [@LupienMcEwenGunnar:2009]. While we did find sex-specific evidence of an effect, their clinical relevance is questionable.

Our findings should be interpreted in light of the following limitations and strengths. Limitations include the cross-sectional design of the present study. Importantly, the non-persistent \acr{edcs} were measured in a pool of night and morning urine samples before the clinical visit, to represent exposure over the previous day, whereas the glucocorticosteroids were measured in the night urine sample. Although we included a wide range of confounders there is the possibility, as with other observational studies, of residual confounding, which might lead to a bias away from the null. Some of the confounders indicated in the adjustment sets had to be remove due to large fractions of missing values. There is further the possibility of misspecification of the outcome model, although we included a spline of the exposure to relax some of the linearity assumptions. The use of more data-adaptive learners was excluded due to the relatively small sample size. We finally acknowledge the possibility that some of chemicals might not act independently (mixture effect). Further research is thus warranted.

@@

In conclusion, in a study of 1,297 children from 6 European birth cohorts, we observed that (i) exposure to non-persistent \acr{edcs} in childhood might have short-term effects on \acr{hitrtse} in childhood, (ii) exposure to non-persistent \acr{edcs} in childhood might disrupt the \acr{hpa} axis in childhood, and (iii) disruption of the \acr{hpa} axis in childhood might have short-term, sex-specific effects on \acr{hitrtse}. Future studies should investigate how glucocorticosteroids might mediate the adverse effects of exposure to \acr{edcs} on childhood neurodevelopment in larger populations.
